---
title: "Data 308: Lecture 6"
subtitle: "Conditions for inference with regression (ISL 3.3.3 )"
author: "CSUCI - Dr. Catalina Medina"
format:
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
    auto-stretch: false
echo: false
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(broom)
library(patchwork)

theme_set(theme_bw(base_size = 24))

elephants <- read_csv(
  "http://csuci-data308-fa26.github.io/course-materials/data/elephants.csv",
  show_col_types = FALSE
) |>
  janitor::clean_names()
```

## Learning goals

By the end of this lecture, you should be able to:

State the conditions needed for inference with linear regression.

Use study design and diagnostic plots to assess whether those conditions are reasonable.

Explain how violations of each condition can affect coefficient estimates, standard errors, confidence intervals, and hypothesis tests.

Recommend possible responses when model conditions are not satisfied.

##

We discussed how simple linear regression can be used to solve different goals

**Prediction** 

- We want the fitted line so we can use it on future data

**Inference**

- Estimate the association between $Y$ and $X$ (confidence interval)
- Test if there is a linear association between $Y$ and $X$ (hypothesis test)


## 

Recall our linear regression from last week
```{r}
ggplot(elephants, aes(x = height, y = mass)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Height (cm)", y = "Mass (kg)")
```

. . .

*Was this model appropriate?*

## Why would we care if the model is "appropriate"?

::::{.columns}
:::{.column width="50%"}
```{r}
#| echo: false
ggplot(elephants, aes(x = height, y = mass)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Height (cm)", y = "Mass (kg)")
```

:::

:::{.column width="50%"}

- it may be chance that we got a good fit for our sample, but the model might be a bad fit for the population
- our estimates relied on our conditions, so we need to understand the implications of unsatisfied conditions

:::
::::

## Conditions for SLR

$$Y_i = \beta_0 + \beta_1 X_i + \epsilon_i$$

1. Linearity: Linear relationship between $X$ and $Y$ ($E(\epsilon_i|X_i) = 0)$.
3. Independence: Independence of $\epsilon_i$.
2. Homoskedasticity: Constant variance of $\epsilon_i$ ($Var(\epsilon_i|X_i) = \sigma^2)$.
4. Normality: Normally distributed $\epsilon_i|X_i$.



# Checking model conditions

:::{.callout-note}
Our four conditions are conditions on the population errors, which we cannot observe. We use the **sample residuals** to look for evidence that the conditions are / are not satisfied.
:::

## Visual checks versus statistical tests

Formal tests of model conditions exist, their performance can depend on sample size and do not indicate the severity of a condition violation.

We will use diagnostic plots together with knowledge of how the data were collected.

- Visual assessment involves judgment.
- We should show our diagnostic plots and explain the evidence behind our decisions.


## Diagnosing linearity

Examine plot of residuals vs fitted:

- Residuals should be scattered around zero without a discernible pattern.

```{r}
set.seed(20250929)

num_samples <- 75

sim_data_linearity_good <- data.frame(x = seq(-2, 2, by = 0.01)) |> 
    mutate(
      y = 5 + 4*x + rnorm(length(x), 0 , 1.5),
      fit = "Good"
    ) |> 
    sample_n(num_samples)

sim_data_linearity_bad <- data.frame(x = seq(-2, 2, by = 0.01)) |> 
    mutate(
      y = 5 + x + 3*x^2 + rnorm(length(x), 0 , 2),
      fit = "Concerning"
    ) |> 
    sample_n(num_samples)


model_linearity_good <- lm(y ~ x, data = sim_data_linearity_good)
model_linearity_bad <- lm(y ~ x, data = sim_data_linearity_bad)

sim_data_linearity <- bind_rows(
  sim_data_linearity_good |>
    modelr::add_predictions(model_linearity_good) |> 
    modelr::add_residuals(model_linearity_good),
  sim_data_linearity_bad |>
    modelr::add_predictions(model_linearity_bad) |> 
    modelr::add_residuals(model_linearity_bad)
)

```


```{r  echo = FALSE, message = FALSE}
ggplot(sim_data_linearity, aes(x = pred, y = resid)) +
  geom_point() +
  geom_smooth(
    method = "loess",
    formula = y ~ x,
    se = FALSE,
    color = "blue"
  ) +
  geom_hline(
    yintercept = 0,
    color = "gray50",
    linetype = "dashed"
  ) +
  facet_grid(cols = vars(fit), scales = "free_x") +
  labs(
    x = "Fitted value",
    y = "Residual"
  )
```

## When linearity is violated

Implications: 

- poor fit
- possibly biased inference

Remedy:

- consider transformations of $X$ or other relationships

## Diagnosing Independence

- Independence generally cannot be established from the standard residual plots.
- Consider how the data were collected (simple random sample? no repeat sampling?)

## When independence is violated

Implications:

- Does not inherently bias our estimators $\hat{\beta}_0$ and $\hat{\beta}_1$
- Our estimates of $SE(\hat{\beta}_p)$ is unsupported (typically underestimate) ($p=0,1$)
- Increased type one error rates (false positive)

Remedy:

- Consider a different model that accounts for correlation (e.g. if you have longitudinal data then a linear mixed effects model may be appropriate)

## Diagnosing homoskedasticity

Examine plot of residuals vs fitted:

- The spread of the residuals should be roughly constant (no fanning or funnel shape)

```{r}
sim_data_variance_good <- data.frame(x = seq(-2, 2, by = 0.01)) |> 
    mutate(
      y = 5 + 2 * x + rnorm(length(x), 0 , 1.5),
      fit = "Good"
    ) |> 
    sample_n(num_samples)

sim_data_variance_bad <- data.frame(x = seq(-2, 2, by = 0.01)) |> 
    mutate(
      y = 5 + 2 * x + rnorm(length(x), mean = 0, sd = abs(x) / 0.5),
      fit = "Concerning"
    ) |> 
    sample_n(num_samples)


model_variance_good <- lm(y ~ x, data = sim_data_variance_good)
model_variance_bad <- lm(y ~ x, data = sim_data_variance_bad)

sim_data_variance <- bind_rows(
  sim_data_variance_good |>
    modelr::add_predictions(model_variance_good) |> 
    modelr::add_residuals(model_variance_good),
  sim_data_variance_bad |>
    modelr::add_predictions(model_variance_bad) |> 
    modelr::add_residuals(model_variance_bad)
)

```


```{r  echo = FALSE, message = FALSE}
ggplot(sim_data_variance, aes(x = pred, y = resid)) +
  geom_point() +
  geom_smooth(
    method = "loess",
    formula = y ~ x,
    se = FALSE,
    color = "blue"
  ) +
  geom_hline(
    yintercept = 0,
    color = "gray50",
    linetype = "dashed"
  ) +
  facet_grid(cols = vars(fit), scales = "free_x") +
  labs(
    x = "Fitted value",
    y = "Residual"
  )
```


## When homoskedasticity is violated

Implications:

- Does not inherently bias our estimators $\hat{\beta}_0$ and $\hat{\beta}_1$
- Our estimates of $SE(\hat{\beta}_p)$ are unsupported ($p=0, 1$)

Remedies:

- Use heteroskedasticity-robust standard errors when the mean
  model is still appropriate.
- Consider a scientifically meaningful transformation of $Y$.
- Use weighted least squares if the variance pattern can be modeled.

## Diagnosing normality: standardized residuals

Examine a Q-Q plot or histogram of standardized residuals ($r_i$)

$$r_i = \frac{e_i - 0}{\text{estimated SD of }e_i}$$
- Standardizing (subtract mean & divide by sd) tells you magnitude of a difference
- The standardized residuals should *roughly* follow the reference line, without systematic curvature or substantial departures
- Most $r_i$ values should be near zero with only a small proportion $|r_i| > 3$.

## Diagnosing normality

```{r}
sim_data_norm_good <- data.frame(x = seq(-2, 2, by = 0.01)) |> 
    mutate(
      y = 5 + 2 * x + rnorm(length(x), mean = 0, sd = 1.5),
      fit = "Good"
    ) |> 
    sample_n(num_samples)

sim_data_norm_bad <- data.frame(x = seq(-2, 2, by = 0.01)) |> 
    mutate(
      y = 5 + 2 * x + rexp(length(x)) - 1,
      fit = "Concerning"
    ) |> 
    sample_n(num_samples)


model_norm_good <- lm(y ~ x, data = sim_data_norm_good)
model_norm_bad <- lm(y ~ x, data = sim_data_norm_bad)

sim_data_norm <- bind_rows(
  sim_data_norm_good |>
    modelr::add_predictions(model_norm_good) |> 
    modelr::add_residuals(model_norm_good),
  sim_data_norm_bad |>
    modelr::add_predictions(model_norm_bad) |> 
    modelr::add_residuals(model_norm_bad)
) |> 
  mutate(fit = factor(fit, levels = c("Good", "Concerning")))
```


```{r }
ggplot(sim_data_norm, aes(sample = resid)) + 
  stat_qq() + 
  stat_qq_line(color = "blue") +
  facet_grid(cols = vars(fit)) +
  labs(
    x = "Theoretical quantiles",
    y = "Residual"
  )
```

## When normality is violated

Implications:

- With a large sample the sampling distributions of $\hat{\beta}_p$ ($p=0,1$) are approximately normally distributed
  - Prediction intervals may have incorrect coverage because their calculation relies on the assumed error distribution
- With a small sample, our estimates of $SE(\hat{\beta}_p)$ are unsupported

Remedies:

- Consider a meaningful transformation of $Y$
- Consider bootstrap confidence intervals

## Visual checks in practice

The `autoplot()` function from the `ggfortify` R package can produce the plots we want for assessing model fit.

| Plot | Primary purpose |
|:--------|:------------------------|
| Residuals vs. fitted | Nonlinearity and nonconstant variance |
| Normal Q-Q | Distribution of errors |
| Scale-location | Nonconstant variance |
| Residuals vs. leverage | Potentially influential observations |



## Assessing conditions for our model

```{r}
#| fig-align: center
#| output-location: slide
#| echo: true
#| out-width: 100%
#| out-height: 100%
#| fig-width: 14
#| fig-height: 9
model_elephant <- lm(mass ~ height, data = elephants)

library(ggfortify)

autoplot(model_elephant)
```

# Conditions  vs model fitting issues

The two lists differ because they answer different questions:

- Model conditions describe the population data-generating process under which particular regression procedures have their advertised properties (e.g., bias and standard errors). 
- Extreme observations are features of an observed dataset that can make the fitted model inaccurate, unstable, or difficult to interpret.

# Extreme observations

> How dependent is the fitted model (and therefore conclusions) on the particular observations in our sample?

Extreme observations can have large residuals and/or unusual x-values

## Identifying influential points

```{r}
#| out-width: 100%
#| out-height: 100%
#| fig-width: 14
#| fig-height: 9
#| fig-align: center
set.seed(20250929)

baseline_data <- tibble(
  x = seq(0, 10, length.out = 40),
  y = 5 + 0.75 * x + rnorm(40, sd = 1.25),
  observation = "Original observation"
)

baseline_model <- lm(y ~ x, data = baseline_data)

comparison_data <- bind_rows(
  baseline_data |>
    mutate(panel = "Baseline"),

  # A: influential, high leverage, and an outlier
  bind_rows(
    baseline_data,
    tibble(
      x = 16,
      y = as.numeric(
        predict(baseline_model, newdata = tibble(x = 16)) - 12
      ),
      observation = "Added observation"
    )
  ) |>
    mutate(panel = "A"),

  # B: outlier, but neither high leverage nor very influential
  bind_rows(
    baseline_data,
    tibble(
      x = 5,
      y = as.numeric(
        predict(baseline_model, newdata = tibble(x = 5)) + 3.5
      ),
      observation = "Added observation"
    )
  ) |>
    mutate(panel = "B"),

  # C: high leverage, but not an outlier or influential
  bind_rows(
    baseline_data,
    tibble(
      x = 16,
      y = as.numeric(
        predict(baseline_model, newdata = tibble(x = 16))
      ),
      observation = "Added observation"
    )
  ) |>
    mutate(panel = "C"),

  # D: influential and high leverage, but not an outlier
  bind_rows(
    baseline_data,
    tibble(
      x = 20,
      y = as.numeric(
        predict(baseline_model, newdata = tibble(x = 20)) - 3
      ),
      observation = "Added observation"
    )
  ) |>
    mutate(panel = "D"),

  # E: influential and an outlier, but not high leverage
  bind_rows(
    baseline_data,
    tibble(
      x = 0,
      y = as.numeric(
        predict(baseline_model, newdata = tibble(x = 0)) + 15
      ),
      observation = "Added observation"
    )
  ) |>
    mutate(panel = "E")
) |>
  mutate(
    panel = factor(
      panel,
      levels = c("Baseline", "A", "B", "C", "D", "E")
    )
  )

model_a <- lm(y ~ x, data = filter(comparison_data, panel == "A"))
model_b <- lm(y ~ x, data = filter(comparison_data, panel == "B"))
model_c <- lm(y ~ x, data = filter(comparison_data, panel == "C"))
model_d <- lm(y ~ x, data = filter(comparison_data, panel == "D"))
model_e <- lm(y ~ x, data = filter(comparison_data, panel == "E"))

model_coeffs <- data.frame(
  panel = c("Baseline", "A", "B", "C", "D", "E"),
  beta0_hat = c(NA, coef(model_a)[1], coef(model_b)[1], coef(model_c)[1], coef(model_d)[1], coef(model_e)[1]),
  beta1_hat = c(NA, coef(model_a)[2], coef(model_b)[2], coef(model_c)[2], coef(model_d)[2], coef(model_e)[2])
) |>
  mutate(
    panel = factor(
      panel,
      levels = c("Baseline", "A", "B", "C", "D", "E")
    )
  )

ggplot(comparison_data, aes(x = x, y = y)) +
  # Model fitted to the baseline observations
  geom_abline(
    intercept = coef(baseline_model)[1],
    slope = coef(baseline_model)[2],
    color = "gray50",
    linetype = "dashed",
    linewidth = 1
  ) +
  # Model fitted to all observations within each panel
  geom_abline(
    data = model_coeffs,
    mapping = aes(intercept = beta0_hat, slope = beta1_hat),
    color = "blue",
    linewidth = 1
  ) +
  geom_point(
    aes(
      color = observation,
      shape = observation
    ),
    size = 2.5
  ) +
  facet_wrap(vars(panel), ncol = 2) +
  scale_color_manual(
    values = c(
      "Original observation" = "gray40",
      "Added observation" = "#D55E00"
    )
  ) +
  scale_shape_manual(
    values = c(
      "Original observation" = 16,
      "Added observation" = 17
    )
  ) +
  labs(
    x = "X",
    y = "Y",
    color = NULL,
    shape = NULL
  ) +
  theme(
    legend.position = "none"
  )
```

## Extreme observations in the diagnostic plot

**What about in higher dimensions?** The residuals-versus-leverage plot helps us identify observations that may strongly influence the fitted model.

- Large leverage indicates high potential to pull the line.
- Large standardized residuals indicate large errors.
- Look into these extreme observations: Does the fitted model appear sensitive to this observation?

## 

```{r}
#| out-width: 100%
#| out-height: 100%
#| fig-width: 14
#| fig-height: 9
#| fig-align: center
plot_baseline <- autoplot(baseline_model, 5)@plots[[1]]
plot_a <- autoplot(model_a, 5)@plots[[1]] + labs(title = "A")
plot_b <- autoplot(model_b, 5)@plots[[1]] + labs(title = "B")
plot_c <- autoplot(model_c, 5)@plots[[1]] + labs(title = "C")
plot_d <- autoplot(model_d, 5)@plots[[1]] + labs(title = "D")
plot_e <- autoplot(model_e, 5)@plots[[1]] + labs(title = "E")

plot_baseline +
  plot_a +
  plot_b +
  plot_c +
  plot_d +
  plot_e +
  plot_layout(
    ncol = 2,
    axis_titles = "collect"
  )
```




## A workflow for assessing your model

Before examining diagnostic plots:

1. Use the study design to assess whether observations are independent. (**independence check**)

After fitting the model:

2. Examine residuals versus fitted values plot:
   - Is there systematic curvature? (**linearity check**)
   - Does the residual spread change? (**constant variance check**)

## A workflow for assessing your model

3. Examine the normal Q-Q plot:
   - Are there strong departures from normality? (**normality check**)

4. Examine residuals versus leverage:
   - Might one or a few observations strongly affect the results? (**model sensitivity**)

## If you identify a concern

Investigate its cause, consider an appropriate remedy, and report
how it affects your conclusions.

**Do not** automatically remove observations or reject a model because a diagnostic plot is imperfect.
