---
title: "Data 308: Lecture 6a"
subtitle: "Conditions for inference with regression (ISL 3.3.3 )"
author: "CSUCI - Dr. Catalina Medina"
format:
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
    auto-stretch: false
echo: false
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(broom)

theme_set(theme_bw(base_size = 24))

elephants <- read_csv(
  "http://csuci-data308-fa26.github.io/course-materials/data/elephants.csv",
  show_col_types = FALSE
) |>
  janitor::clean_names()
```

## Learning goals

By the end of this lecture, you should be able to:

State the conditions needed for inference with linear regression.

Use study design and diagnostic plots to assess whether those conditions are reasonable.

Explain how violations of each condition can affect coefficient estimates, standard errors, confidence intervals, and hypothesis tests.

Recommend possible responses when model conditions are not satisfied.

##

We discussed how simple linear regression can be used to solve different goals

**Prediction** 

- We want the fitted line so we can use it on future data

**Inference**

- Estimate the association between $Y$ and $X$ (confidence interval)
- Test if there is a linear association between $Y$ and $X$ (hypothesis test)


## 

Recall our linear regression from last week
```{r}
ggplot(elephants, aes(x = height, y = mass)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Height (cm)", y = "Mass (kg)")
```

. . .

*Was this model appropriate?*

## Why would we care if the model is "appropriate"?

::::{.columns}
:::{.column width="50%"}
```{r}
#| echo: false
ggplot(elephants, aes(x = height, y = mass)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Height (cm)", y = "Mass (kg)")
```

:::

:::{.column width="50%"}

- it may be chance that we got a good fit for our sample, but the model might be a bad fit for the population
- our estimates relied on our conditions, so we need to understand the implications of unsatisfied conditions

:::
::::

## Conditions for SLR

$$Y_i = \beta_0 + \beta_1 X_i + \epsilon_i$$

1. Linearity: Linear relationship between $X$ and $Y$ ($E(\epsilon_i|X_i) = 0)$.
3. Independence: Independence of $\epsilon_i$.
2. Homoskedasticity: Constant variance of $\epsilon_i$ ($Var(\epsilon_i|X_i) = \sigma^2)$.
4. Normality: Normally distributed $\epsilon_i|X_i$.



# Checking model conditions

:::{.callout-note}
Our four conditions are conditions on the population errors, which we cannot observe. We use the **sample residuals** to look for evidence that the conditions are / are not satisfied.
:::

## Visual checks versus statistical tests

Formal tests of model conditions exist, their performance can depend on sample size and do not indicate the severity of a condition violation.

We will use diagnostic plots together with knowledge of how the data were collected.

- Visual assessment involves judgment.
- We should show our diagnostic plots and explain the evidence behind our decisions.


## Diagnosing linearity

Examine plot of residuals vs fitted:

- Residuals should be scattered around zero without a discernible pattern.

```{r}
set.seed(20250929)

num_samples <- 75

sim_data_linearity_good <- data.frame(x = seq(-2, 2, by = 0.01)) |> 
    mutate(
      y = 5 + 4*x + rnorm(length(x), 0 , 1.5),
      fit = "Linear"
    ) |> 
    sample_n(num_samples)

sim_data_linearity_bad <- data.frame(x = seq(-2, 2, by = 0.01)) |> 
    mutate(
      y = 5 + x + 3*x^2 + rnorm(length(x), 0 , 2),
      fit = "Nonlinear"
    ) |> 
    sample_n(num_samples)


model_linearity_good <- lm(y ~ x, data = sim_data_linearity_good)
model_linearity_bad <- lm(y ~ x, data = sim_data_linearity_bad)

sim_data_linearity <- bind_rows(
  sim_data_linearity_good |>
    modelr::add_predictions(model_linearity_good) |> 
    modelr::add_residuals(model_linearity_good),
  sim_data_linearity_bad |>
    modelr::add_predictions(model_linearity_bad) |> 
    modelr::add_residuals(model_linearity_bad)
)

```


```{r  echo = FALSE, message = FALSE}
ggplot(sim_data_linearity, aes(x = pred, y = resid)) +
  geom_point() +
  geom_smooth(
    method = "loess",
    formula = y ~ x,
    se = FALSE,
    color = "blue"
  ) +
  geom_hline(
    yintercept = 0,
    color = "gray50",
    linetype = "dashed"
  ) +
  facet_grid(cols = vars(fit), scales = "free_x") +
  labs(
    x = "Fitted value",
    y = "Residual"
  )
```

## When linearity is violated

Implications: 

- poor fit
- possibly biased inference

Remedy:

- consider transformations of $X$ or other relationships

## Diagnosing Independence

- Independence generally cannot be established from the standard residual plots.
- Consider how the data were collected (random sample? no repeat sampling?) and on whom the data were collected (would you expect independence?)
- Dependence could be revealed by plotting residuals versus unmodeled variables (time, class, etc)

## When independence is violated

Implications:

- Does not inherently bias our estimators $\hat{\beta}_0$ and $\hat{\beta}_1$
- Our standard errors are likely incorrect (typically underestimate)
- Increased type one error rates (false positive)

Remedy:

- Consider a different model that accounts for correlation (e.g. if you have longitudinal data then a linear mixed effects model may be appropriate)

## Diagnosing homoskedasticity

Examine plot of residuals vs fitted:

- The spread of the residuals should be roughly constant (no fanning or funnel shape)

```{r}
sim_data_variance_good <- data.frame(x = seq(-2, 2, by = 0.01)) |> 
    mutate(
      y = 5 + 2 * x + rnorm(length(x), 0 , 1.5),
      fit = "Constant variance"
    ) |> 
    sample_n(num_samples)

sim_data_variance_bad <- data.frame(x = seq(-2, 2, by = 0.01)) |> 
    mutate(
      y = 5 + 2 * x + rnorm(length(x), mean = 0, sd = abs(x) / 0.5),
      fit = "Nonconstant variance"
    ) |> 
    sample_n(num_samples)


model_variance_good <- lm(y ~ x, data = sim_data_variance_good)
model_variance_bad <- lm(y ~ x, data = sim_data_variance_bad)

sim_data_variance <- bind_rows(
  sim_data_variance_good |>
    modelr::add_predictions(model_variance_good) |> 
    modelr::add_residuals(model_variance_good),
  sim_data_variance_bad |>
    modelr::add_predictions(model_variance_bad) |> 
    modelr::add_residuals(model_variance_bad)
)

```


```{r  echo = FALSE, message = FALSE}
ggplot(sim_data_variance, aes(x = pred, y = resid)) +
  geom_point() +
  geom_smooth(
    method = "loess",
    formula = y ~ x,
    se = FALSE,
    color = "blue"
  ) +
  geom_hline(
    yintercept = 0,
    color = "gray50",
    linetype = "dashed"
  ) +
  facet_grid(cols = vars(fit), scales = "free_x") +
  labs(
    x = "Fitted value",
    y = "Residual"
  )
```


## When homoskedasticity is violated

Implications:

- Does not inherently bias our estimators $\hat{\beta}_0$ and $\hat{\beta}_1$
- Our standard errors are likely incorrect

Remedies:

- Use heteroskedasticity-robust standard errors when the mean
  model is still appropriate.
- Consider a scientifically meaningful transformation of $Y$.
- Use weighted least squares if the variance pattern can be modeled.

## Diagnosing normality: standardized residuals

Examine a Q-Q plot or histogram of standardized residuals ($r_i$)

$$r_i = \frac{e_i - 0}{\text{estimated SD of }e_i}$$

- The standardized residuals should *roughly* follow the reference line, without systematic curvature or substantial departures
- Most $r_i$ values should be near zero with only a small proportion $|r_i| > 3$.

## Diagnosing normality

```{r}
sim_data_norm_good <- data.frame(
  resid = rnorm(100, mean = 0, sd = 1.5),
  fit = "Normal"
)
sim_data_norm_bad <- data.frame(
  resid = rexp(100) - 1,
  fit = "Nonnormal"
)

sim_data_norm <- bind_rows(sim_data_norm_good, sim_data_norm_bad) |> 
  mutate(fit = factor(fit, levels = c("Normal", "Nonnormal")))
```


```{r }
ggplot(sim_data_norm, aes(sample = resid)) + 
  stat_qq() + 
  stat_qq_line(color = "blue") +
  facet_grid(cols = vars(fit)) +
  labs(
    x = "Theoretical quantiles",
    y = "Residual"
  )
```

## When normality is violated

Implications:

- With no outliers and a large enough sample the sampling distribution of the coefficient estimator becomes approximately normal
- Can have biased predictions for $Y$

Remedies:

- Consider a scientifically meaningful transformation of $Y$.

## Visual checks in practice

The `autoplot()` function from the `ggfortify` R package can produce the plots we want for assessing model fit.
Recall residuals: $e_i = y_i - \hat{y}_i$


| Plot | Primary purpose |
|:--------|:------------------------|
| Residuals vs. fitted | Nonlinearity and nonconstant variance |
| Normal Q-Q | Distribution of errors |
| Scale-location | Nonconstant variance |
| Residuals vs. leverage | Potentially influential observations |

## Assessing conditions for our model

```{r}
#| fig-align: center
#| output-location: slide
#| echo: true
model_elephant <- lm(mass ~ height, data = elephants)

library(ggfortify)

autoplot(model_elephant)
```

<!-- ## Unusual and influential observations -->

<!-- The four conditions describe the behavior of the errors across the population. Influence instead asks whether our fitted results depend too heavily on one or a few observations. -->

<!-- - An **outlier** has an unusually large residual. -->
<!-- - A **high-leverage observation** has an unusual value of $X$. -->
<!-- - An **influential observation** substantially changes the fitted -->
<!--   model when included or removed. -->

<!-- An observation can have high leverage without being influential, and an outlier does not necessarily have high leverage. -->