---
title: "DATA 208: Lab 1a"
subtitle: "Familiarizing yourself with a Quarto document"
date: "2026-08-25"
author: "Your name here"
lang: en
format: 
  html:
    echo: false
    message: false
    warning: false
    toc: true
    toc-depth: 4
    embed-resources: true
---

```{r}
library(tidyverse)
```

Start off this lab by clicking the "Render" (the blue arrow) at the top. This will produce an html report, where you can view the plots and questions. The .qmd file is a Quarto document, this document is where you will make changes, render, and then see those changes reflected in the html report it procudes.


In this lab we will be exploring California unemployment rates.
This data is available on the [CA Open Data Portal](https://data.ca.gov/dataset/unemployment-rate-by-age-groups).

We will begin by reading in this data and saving it into an **object** called `unemployment`.

```{r}
unemployment <- read_csv("http://csuci-data308-fa26.github.io/course-materials/week-01/data/unemployment_rate_by_age_groups.csv")
```

We can view what is in the `employment` object by printing it.

```{r}
unemployment
```

You can skip over the following code chunk, I needed to do some data cleaning and wrangling to get it in a good format for plotting.

```{r}
unemployment_new <- unemployment |>
  mutate(new_date = as_date(Date, format = "%m/%d/%Y")) |> 
  pivot_longer(
    cols = starts_with("Age "),
    names_to = "age_group",
    names_prefix = "Age ",
    values_to = "unemployment_rate"
  )
```


## Unemployment over time

```{r}
unemployment_new |> 
  filter(age_group == "25-34") |> 
  ggplot(aes(x = new_date, y = unemployment_rate)) +
  geom_point() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    x = "Date",
    y = "Unemployment rate \n for individuals ages 25-34"
  )
```

## Question 1

If you were to describe these trends to someone what would you say?

### Response

[You can erase this and type your response here, without the brackets.]

## Question 2

What is a "typical" unemployment rate? What would be an unusual rate?

### Response

[You can erase this and type your response here, without the brackets.]

## Unemployment trends for each age group

```{r}
ggplot(
  unemployment_new, 
  aes(x = new_date, y = unemployment_rate, color = age_group, linetype = age_group)
) +
  geom_line() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    x = "Date",
    y = "Unemployment rate",
    color = "Age group",
    linetype = "Age group"
  )
```

## Question 3

Given this data, what would you predict the average unemployment rate of 2027 to be for people aged 25-34? Why?

### Response

[You can erase this and type your response here, without the brackets.]

## Question 4

Given this data, what would you predict the average unemployment rate of 2035 to be for people aged 55-64? Why?

### Response

[You can erase this and type your response here, without the brackets.]


## Question 5

Which age group(s) had the highest unemployment rate in general?

### Response

[You can erase this and type your response here, without the brackets.]

## Question 6

Which age group(s) had the lowest unemployment rate in general?

### Response

[You can erase this and type your response here, without the brackets.]

## Question 7

Would you say there is a meaningful difference in unemployment rates between 55-64 year olds and 20-24 year olds? Why?

### Response

[You can erase this and type your response here, without the brackets.]

## Question 8

Would you say there is a meaningful difference in unemployment rates between 55-64 year olds and 35-34 year olds? Why?

### Response

[You can erase this and type your response here, without the brackets.]


When you are done, click "Render" (the blue arrow) at the top. Submit the rendered html report to CI Learn.
