---
title: "DATA 308: Lab 1b"
subtitle: "Working with data"
date: "2026-08-27"
author: "Your name here"
lang: en
format:
  pdf:
    pdf-standard: [a-2b, ua-2] # Configures PDF/A and PDF/UA-2 compliance
    pdf-engine: lualatex       # LuaLaTeX is highly recommended for tagging support
---

When working in a Quarto document I recommend:

- Start by rendering your document, so you know if there are any issues with the file.
- Render periodically as you work and inspect your rendered file, so you know your document and code are behaving as desired.

## Getting your data

0. In the code chunk below load in the `tidyverse` package.

```{r}

```


1. In the code chunk below read in your data and save it into an object called `game_films`.

```{r}

```


2. Use the `glimpse()` function to inspect your data. How many rows and columns are there? Note, I did not add a code chunk for you. You will first need to click the green "+C" button at the top right of this document to add a code chunk.


## Explore your data

3. Reproduce a plot from the lecture slides.


4. Create a box plot using `geom_boxplot()` for a numerical variable of your choice.


5. Create a box plot of `worldwide_box_office` by `worldwide_box_office_currency`. Hint: you can do this by specifying `worldwide_box_office_currency` for your y-axis.


6. Create a scatterplot for two numerical variables of your choosing, and add a smoothed curve.


7. Calculate the minimum, maximum, 25\% quantile, and 75\% quantile of `rotten_tomatoes` scores.


## Version control

I do recommend using GitHub for this class, but I will not require for homework assignments, just the final project.
I recommend you practice by using a single GitHub repository for your labs and homework assignments for this class.

Here are [instructions I created for installing and configuring GitHub and RStudio]( https://csuci-data-200-f26.github.io/lecture-materials/week-00/software-setup). (Skip step 4)

