---
title: "Data 308: Lecture 5 (5a & 5b)"
subtitle: "Hypothesis testing for a regression slope"
author: "CSUCI - Dr. Catalina Medina"
format:
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
    auto-stretch: false
echo: false
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(broom)

theme_set(theme_bw(base_size = 24))

elephants <- read_csv(
  "http://csuci-data308-fa26.github.io/course-materials/data/elephants.csv",
  show_col_types = FALSE
) |>
  janitor::clean_names()

elephant_model <- lm(mass ~ height, data = elephants)
elephant_results <- tidy(elephant_model, conf.int = TRUE)
elephant_slope <- elephant_results |> filter(term == "height")

set.seed(308)
null_n <- 50
null_sigma <- 3
null_samples <- map_dfr(1:1000, function(sample_id) {
  sample_data <- tibble(
    x = runif(null_n, -2, 2),
    y = 2 + rnorm(null_n, 0, null_sigma)
  )

  tibble(
    sample_id = sample_id,
    slope = coef(lm(y ~ x, data = sample_data))["x"]
  )
})

null_examples <- map_dfr(1:4, function(sample_id) {
  set.seed(308 + sample_id)
  tibble(
    sample_id = paste("Sample", sample_id),
    x = runif(null_n, -2, 2),
    y = 2 + rnorm(null_n, 0, null_sigma)
  )
})

null_example_slopes <- null_examples |>
  group_by(sample_id) |>
  group_modify(~ tidy(lm(y ~ x, data = .x))) |>
  filter(term == "x")

relationship_data <- bind_rows(
  tibble(
    relationship = "A",
    x = runif(40, -5, 5),
    y = 2 + 2 * x + rnorm(40, 0, 1.5)
  ),
  tibble(
    relationship = "B",
    x = runif(40, -5, 5),
    y = 2 + -2 * x + rnorm(40, 0, 5)
  ),
  tibble(
    relationship = "C",
    x = runif(40, -5, 5),
    y = rnorm(40, 0, 5.5)
  ),
  tibble(
    relationship = "D",
    x = runif(40, -5, 5),
    y = x^2 + rnorm(40, 0, 1.5)
  )
) |>
  mutate(relationship = factor(relationship, levels = c("A", "B", "C", "D")))

observed_slope_example <- 0.75

null_samples <- null_samples |>
  mutate(
    at_least_as_extreme =
      abs(slope) >= abs(observed_slope_example)
  )

simulated_p_value <- mean(
  null_samples$at_least_as_extreme
)
```

# Goals for this week

- theoretical understanding of the key aspects of a hypothesis test
- ability to conduct a hypothesis test and make a conclusion in context

## Theory-based goals

By the end of this week students should be able to:

- explain what a p-value is
- distinguish between statistical significance and practical significance
- explain Type I and Type II errors conceptually
- identify factors that affect statistical power
- connect a two-sided hypothesis test with a confidence interval.

## Application-based goals

By the end of this week students should be able to:

- conduct a hypothesis test for the slope of simple linear regression:
  - state null and alternative hypotheses (in symbols and in context)
  - fit the model and obtain the p-value in R
  - make a conclusion in context based on the results


# From estimation to testing

## What did we do last week?

- We used a sample to estimate an unknown population slope, $\beta_1$.
- Different random samples produced different estimates, $\hat{\beta}_1$.
- A confidence interval described a range of plausible values for $\beta_1$.

. . .

> What if our research question is specifically whether there is any linear association in the population?

## Two inferential questions

::::{.columns}

:::{.column width="48%"}

### Estimation

> What values of $\beta_1$ are plausible?

Use a confidence interval.

:::

:::{.column width="48%"}

### Hypothesis testing

> Are the data compatible with a particular proposed value of $\beta_1$?

Use a hypothesis test.

:::

::::

## Slopes

For each plot, describe (1) the strength of the relationship, (2) the strength of the linear relationship, (3) visually approximate the slope

```{r}
#| echo: false
#| fig-align: center

ggplot(relationship_data, aes(x, y)) +
  geom_point(size = 2.2, alpha = 0.8) +
  #lims(y = c(-10, 10)) +
  facet_wrap(~ relationship, nrow = 2, scales = "free_x") +
  labs(x = "X", y = "Y")
```

## Slopes (Solution)

| Plot | strength of relationship | strength of linear relationship | slope |
|:---|:------------|:------------|:---|
| A | very strong | very strong | 2 |
| B | strong | weak/moderate | -2 |
| C | no relationship | no linear relationship | 0 |
| D | very strong | no linear relationship | 0 |

* I did slightly edit B to make it a weaker relationship than what we saw in class 5a

# Hypotheses for a regression slope

## Research question

> Is elephant height linearly associated with elephant mass in the population represented by these data?

Our population model is

$$\text{Mass}_i = \beta_0 + \beta_1 \text{Height}_i + \epsilon_i.$$

. . .

Our question is whether $\beta_1$ differs from zero.

## Null hypothesis

In symbols:

$$H_0: \beta_1 = 0$$

In context:

> In the population, there is no linear association between elephant height and mean mass.

. . .

If $\beta_1=0$, the mean of $Y$ does not change linearly with $X$.

## Idea of a hypothesis test

$$H_0: \beta_1 = 0$$

For a hypothesis test, we operate by assuming the null hypothesis is true, then we see if there is evidence against it.

- Is it unlikely to get this data if $H_0$ is true?


## Alternative hypothesis

$$H_A: \beta_1 \neq 0$$

In context:

> In the population, elephant height is linearly associated with mean mass.

. . .

If we find evidence against $H_0$,

- meaning evidence that it is unlikely to get this data if $H_0$ is true,
- then we found evidence in support of $H_A$

## What does our sample show?

```{r}
#| echo: false
#| fig-align: center
#| fig-height: 4.5

ggplot(elephants, aes(x = height, y = mass)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "#B22222") +
  labs(x = "Height (cm)", y = "Mass (kg)")
```

## The sample slope is not zero

```{r}
#| echo: true

coef(elephant_model)
```

> Is a nonzero sample slope enough to reject $H_0$?

. . .

No. Random samples can produce nonzero estimates even when the population slope is zero.

- Recall the sampling distribution of $\hat{\beta}_1$ from lecture 4b

. . .

Then how do we know if we have evidence against $H_0: \beta_1 = 0$?

# Imagine the null hypothesis is true

## A population with no linear association

Suppose the true population model is

$$Y_i = 2 + 0X_i + \epsilon_i.$$

- The population slope is zero.
- $X$ provides no information about the mean of $Y$.
- A random sample can still show an apparent positive or negative slope.

## Four samples from the "null world"

```{r}
#| echo: false
#| fig-align: center
#| fig-height: 5

ggplot(null_examples, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "#B22222") +
  facet_wrap(~ sample_id, nrow = 2) +
  labs(x = "X", y = "Y") +
  geom_label(
    data = null_example_slopes,
    aes(
      x = Inf,
      y = Inf,
      label = paste0(
        "Estimated slope = ",
        round(estimate, 2)
      )
    ),
    inherit.aes = FALSE,
    hjust = 1.05,
    vjust = 1.15,
    size = 4
  )
```

## Many samples from the null world

```{r}
#| echo: false
#| fig-align: center
#| fig-height: 4.5

ggplot(null_samples, aes(x = slope)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "grey80") +
  geom_vline(xintercept = 0, color = "#294957", linewidth = 1.5) +
  labs(
    x = expression("Estimated slope " * hat(beta)[1]),
    y = "Number of samples"
  )
```

## The central question

> If the population slope were zero, how often would random sampling produce a result at least as extreme as ours?

. . .

This question leads to the **p-value**.

# The p-value

## Test statistic

A slope of 5 could be surprising in one study but unsurprising in another.

We compare the estimate with its sampling variability:

$$t = \frac{\hat{\beta}_1 - 0}{SE(\hat{\beta}_1)}.$$

. . .

> The test statistic tells us how many standard errors the estimated slope is from the null value.

## Two estimates of 5

::::{.columns}

:::{.column width="48%"}

$$\hat\beta_1=5,\qquad SE(\hat\beta_1)=1$$

$$t=5$$

The estimate is five standard errors from zero.

:::

:::{.column width="48%"}

$$\hat\beta_1=5,\qquad SE(\hat\beta_1)=10$$

$$t=0.5$$

The estimate is half a standard error from zero.

:::

::::

. . .

If $H_0$ is true, which test statistic is more surprising?

## Concept of a p-value

Suppose a research team collects one of these samples and estimates $\hat{\beta}_1$. The software also determines how many standard errors the estimate is from zero.

If the population slope were actually zero, what is the probability of obtaining an estimated slope at least this many standard errors from zero?

## Definition of a p-value

> A p-value is the probability, assuming the null hypothesis is true, of obtaining a test statistic at least as extreme as the one observed.

. . .

1. Assume $H_0$ is true.
2. Consider the results that could arise through random sampling.
3. Find how often results at least as extreme as ours occur.

## How do we get a p-value?

```{r}
#| echo: true

summary(elephant_model)
```

. . .

Note `<2e-16` means $< 2\times 10^{-16}$, which is the smallest p-value R will report


## Interpreting the elephant p-value

> If there were no linear association between elephant height and mean mass in the population, the probability of obtaining a test statistic at least as far from zero as the one observed, in either direction, would be less than $< 2\times 10^{-16}$.

## Check your understanding

Which is a correct interpretation of $p-value=0.03$?

A. There is a 3% probability that the null hypothesis is true.

B. There is a 3% probability that the results occurred by chance.

C. If the null hypothesis were true, results at least as extreme as ours would occur in about 3% of comparable studies.

D. There is a 97% probability that the alternative hypothesis is true.

. . .

*Solution: C. The others are all similarly wrong.*

## What a p-value does not tell us

The p-value is **not**:

- the probability that $H_0$ is true;
- the probability that the results occurred "by chance";
- the probability that the finding will replicate;
- the size or importance of the association.

# Decision and conclusion

## Significance level

Before examining the result, researchers select a significance level, $\alpha$ (a cutoff).

A common choice is $\alpha=0.05.$

Decision rule:

- If $p-value < \alpha$, reject $H_0$.
- If $p-value \geq \alpha$, fail to reject $H_0$.

## Why "fail to reject"?

A large p-value says that the data do not provide strong evidence against $H_0$.

It does not establish that $H_0$ is true.

. . .

> Absence of evidence is not evidence of absence.

## Conclusion for the elephants

**Statistical decision:**

Because $p-value < 0.05$, we reject $H_0$.

**Conclusion in context:**

> The results provide evidence of a linear association between elephant height and mean mass in the population (p-value $< 2\times 10^{-16}$).

. . .

... But what is that linear relationship? This analysis is incompelte.

## If the result were not statistically significant

**Statistical decision:**

If  $p-value \geq 0.05$, we would fail to reject $H_0$.

**Conclusion in context:**

> The results do not provide sufficient evidence of a linear association between elephant height and mean mass in the population (p-value = ...).

Avoid saying:

- "We accept the null hypothesis."
- "We proved that there is no association."

# Statistical and practical significance

*This is where we stopped for Lec 5a*

## A statistically significant result

Suppose in a very large study, each additional hour of weekly study time is associated with an estimated increase of **0.08 percentage points** in exam score.

$$\hat\beta_1=0.08,\qquad p<0.001$$

Discuss with a partner:

1. Is the result statistically significant?
2. Is it practically significant?
3. What other information would you want?

## Two different questions

::::{.columns}

:::{.column width="48%"}

### Statistical significance

Is the observed association difficult to explain if $H_0$ is true?

Use the p-value.

:::

:::{.column width="48%"}

### Practical significance

Is the estimated association large enough to matter in context?

Use the estimate, confidence interval, and subject knowledge.

:::

::::

## Sample size matters

With a very large sample, even a tiny association can produce a small p-value.

With a small sample, an important association might not produce a small p-value.

. . .

> A small p-value does not imply a large or important association.

## Why report a confidence interval too?

The p-value helps assess evidence against $H_0$.

The confidence interval communicates:

- the estimated direction;
- plausible effect sizes;
- the precision of the estimate;
- information relevant to practical significance.

# Conducting a hypothesis test

## Complete workflow

1. Identify the population parameter.
2. State $H_0$ and $H_A$ in symbols and context.
3. Fit the regression model. Obtain the estimated slope, confidence interval, and p-value.
5. Compare the p-value with $\alpha$. Make a statistical decision.
7. State the conclusion in context.
8. Consider the magnitude and practical importance of the estimate.

## R workflow

```{r}
#| echo: true

elephant_model <- lm(mass ~ height, data = elephants)

confint(elephant_model)

summary(elephant_model)
```

## A complete report

We tested $H_0:\beta_1=0$ against $H_A:\beta_1\ne0$, where $\beta_1$ represents the population slope relating elephant height to mean mass. Assuming the conditions for inference with simple linear regression have been met:

> We found evidence of a linear association between elephant height and average mass in the population of adult Asian elephants (p-value = $<2 \times 10^{-16}$). We estimate an increase of one cm in height to be associated with an increase in average mass of `r round(elephant_slope$estimate, 1)`kg, between `r round(elephant_slope$conf.low, 1)` to `r round(elephant_slope$conf.high, 1)` with 95% confidence.


## Takeaways

- Hypothesis test for $H_0: \beta_1 = 0$ versus $H_A: \beta_1 \neq 0$ tests for evidence of a linear relationship.
- A p-value measures how surprising the observed evidence would be if $H_0$ were true.
- Statistical significance and practical significance answer different questions.
- A complete conclusion describes the population, variables, evidence, and estimated effect.
