---
title: "Data 308: Lecture 7b"
subtitle: "Multiple linear regression continued"
author: "CSUCI - Dr. Catalina Medina"
format:
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
echo: true
---

```{r}
#| echo: false

library(tidyverse)
theme_set(theme_bw(base_size = 24))
```

# After learning multiple linear regression...

The next natural questions are:

1. Does that change or add any assumptions to our model?
2. What variables should I add as $X's$ in addition to my primary explanatory variable of interest? (Lec 8a)


# 1. Conditions

Multiple linear regression has 5 conditions:

1. Linearity (appropriate functional form)
2. Independence of observations
3. Constant variance of errors
4. Normality of errors
5. **No perfect collinearity**

## Perfect collinearity

Perfect collinearity =  An $X_j$ is a perfect linear combination of another

- e.g., Using SAT Math, SAT Reading, & SAT total (total = math + reading) as 3 $X$'s

Cannot fit model mathematically so statistical software will drop at least one $X$

## High Collinearity

High collinearity = highly correlated $X$'s

- e.g., Suppose we wanted to include SAT Math & SAT Reading as 2 $X$'s


Results in: 

- larger standard errors of $\hat{\beta}$'s for highly correlated $X$'s
- difficulty separating the association between $Y$ and $X_j$ from $Y$ and $X_k$
- mathematical instability resulting in high sensitivity of results to sample

## High Collinearity

Remedy:

- Determine whether both variables are scientifically necessary
- Do not remove an important confounding variable solely because it is highly correlated with another predictor
  - Note: VIF is discussed in the ISL textbook, but we should not remove a variable only because it has a high VIF

# Linear regression as a tool

When we perform multiple linear regression, we usually are interested in answering a few important questions.

1. Is at least one of the predictors $X_1, X_2,...,X_p$ useful in predicting the response?
2. Do all the predictors help to explain $Y$ , or is only a subset of the predictors useful?
3. How well does the model fit the data?
4. Given a set of predictor values, what response value should we predict, and how accurate is our prediction?

# Q1: Is There a Linear Relationship Between the Response and Predictors?

## Q1 with simple linear regression

$$Y_i = \beta_0 + \beta_1 X_i + \epsilon_i$$

. . .

Hypothesis test for $\beta_1$! 

- $H_0: \beta_1 = 0$ & $H_A: \beta_1 \neq 0$
- Results for an F-test are shown when you use `summary()` or `tidy()` on your model
- Make your statistical decision and conclusion in context. Report confidence interval!


## Q1 with multiple linear regression

$$Y_i = \beta_0 + \beta_1 X_{1i}  + \beta_2 X_{2i} + ... + \beta_p X_{p,i} + \epsilon_i$$

. . .

Hypothesis test for all $\beta_j$'s!

. . .

- $H_0:$ All $\beta_j$'s equal 0 & $H_A$: At least one $\beta_j$ is non-zero
- Results for an F-test are shown when you use `summary()` or `glance()` on your model
- Make your statistical decision and conclusion in context. Report confidence interval!

## F-test (in practice)

The dataset contains data on test performance, school characteristics and student demographic backgrounds for school districts in California in 1999.

```{r}
data("CASchools", package = "AER")

ca_schools <- CASchools |>
  mutate(student_teacher_ratio = students / teachers) |>
  mutate(student_computer_ratio = case_when(
    computer != 0 ~ students / computer,
    computer == 0 ~ 0
  )) |>
  mutate(ave_score = (read + math) / 2)
```


## California Schools data

::::{.columns}

:::{.column width="50%"}
```{r}
#| echo: false
ggplot(
  ca_schools,
  aes(x = ave_score)
) +
  geom_histogram() +
  labs(
    x = "Average test score",
    y = "Count"
  )
```

:::

:::{.column width="50%"}

```{r}
#| echo: false
ggplot(
  ca_schools,
  aes(x = student_teacher_ratio, y = ave_score)
) +
  geom_point() +
  labs(
    x = "Student-to-Teacher Ratio",
    y = "Average test score"
  )
```
:::

::::

::::{.columns}

:::{.column width="50%"}

```{r}
#| echo: false
ggplot(
  ca_schools,
  aes(x = student_computer_ratio, y = ave_score)
) +
  geom_point() +
  labs(
    x = "Student-to-Computer Ratio",
    y = "Average test score"
  )
```
:::

:::{.column width="50%"}

```{r}
#| echo: false
ggplot(
  ca_schools,
  aes(x = lunch, y = ave_score)
) +
  geom_point() +
  labs(
    x = "% Qualifying for reduced-price lunch",
    y = "Average test score"
  )
```
:::
::::



## California Schools data

We are going to regress average test score on student-to-teacher ratio, student-to-computer ratio, and percent qualifying for reduced-price lunch.

\begin{equation*}
  \begin{split}
    &\text{Score}_i = \beta_0 \\
    &+ \beta_1 \text{Student-Teacher Ratio}_i\\
    &+ \beta_2 \text{Student-Computer Ratio}_i \\
    &+ \beta_3 \text{\% Reduced Price Lunch}_i\\
    &+ \epsilon_i
  \end{split}
\end{equation*}


## F-test (in practice)

```{r}
score_model1 <- lm(
  ave_score ~ student_teacher_ratio + student_computer_ratio + lunch,
  data = ca_schools
)

summary(score_model1)
```

## F-test (in practice)

What if you want to extract the F-statistic or its p-value?

```{r}
library(broom)

glance(score_model1)
```

. . .

```{r}
glance(score_model1) |>
  pull(statistic)

glance(score_model1) |>
  pull(p.value)
```


## F-test (in practice)

$H_0:$ All $\beta_j$'s equal 0 & $H_A$: At least one $\beta_j$ is non-zero

```{r}
#| echo: false

m1_p_value <- glance(score_model1) |>
  pull(p.value) |> 
  format(scientific = TRUE)
```

. . .

**Statistical Decision**: Our p-value is `r m1_p_value`, which is tiny, for sure less than 0.05, so reject $H_0$

. . .

**Conclusion in context**: We found very strong evidence (p-value = `r m1_p_value`) that at least one of student-teacher ratio, student-computer ratio, or % reduced lunch is linearly related to score.

## Partial F-test for nested models

From the plots it was pretty clear lunch is linearly related to score. What if we only wanted to do a test to see if student-teacher ratio and or student-computer ratio is linearly related to score?

. . .

We can do a partial F-test for nested models!

- $H_0: \beta_1 = \beta_2 = 0$
- $H_A:$ At least one of $\beta_1$ or $\beta_2$ are nonzero.

## F-test with subset of coefficients

```{r}
score_model1 <- lm(
  ave_score ~ student_teacher_ratio + student_computer_ratio + lunch,
  data = ca_schools
)

score_model2 <- lm(
  ave_score ~ lunch,
  data = ca_schools
)

anova(score_model2, score_model1)
```



## F-test with subset of coefficients

```{r}
partial_f_test <- anova(score_model2, score_model1)
partial_f_test_p_value <- partial_f_test$`Pr(>F)`[2] |> 
  format(scientific = TRUE)
```


**Statistical Decision**: Our p-value is `r partial_f_test_p_value`, much smaller than 0.05, so reject  $H_0$.

. . .

**Conclusion in context**: We found very strong evidence (p-value = `r partial_f_test_p_value`) that at least one of student-teacher ratio or student-computer ratio is linearly related to score, when accounting for lunch.



## Why an F-test instead of individual tests?

```{r}
tidy(score_model1)
tidy(score_model2)
```

## Inflated Type I Error rate

Consider fitting a multiple linear regression model where all of the slopes are actually zero. In other words, none of the predictors are truly associated with the response.

- Imagine we collect many independent samples from this population and fit the same model to each sample.
- For each predictor, we perform an individual test using $\alpha=0.05$.
- For each test, there is a 5% probability of Type I error.

## Inflated Type I Error rate

- cont...
- For each test, there is a 5% probability of Type I error.
- Even though all $H_0$'s are true, approximately 5% of the tests for each predictor will produce a p-value below 0.05 by chance.
  - This means that even when none of the predictors are associated with the response, we may incorrectly conclude that some are!
  - As we test more predictors, the probability of at least one false positive increases.

# Q2: Do all the predictors help to explain $Y$, or is only a subset of the predictors useful?

## Statistical Inference

If you are doing statistical inference you need to be conscious not to increase your type I error rate needlessly.

- You should not fit a bunch of models, see what "fits best", then do a hypothesis test ("p-hacking")
- Therefore you should consider ahead of time what predictors should be included based on context knowledge (lec 8a)


## Model selection for prediction

Selecting a model for prediction requires evaluating how well candidate models perform on new observations. 

- *Techniques such as cross-validation, model selection, and regularization are beyond the scope of this course and are covered in Chapters 5 and 6 of ISL textbook and in MATH 398 and MATH 408.*


# Q3 How well does the model fit the data?

## Residual Standard Error (RSE)

$$RSE = \hat{\sigma} = \sqrt{\frac{RSS}{n - 2}} = \sqrt{\frac{1}{n - 2} \sum_{i = 1}^{n} (y_i - \hat{y}_i)^2}$$

- Measured in units of $y$

. . .

$n - 2$ is required to for key statistical properties, but it is often replaced by just $n$

- Using $n$ instead of $n - 2$ is called **Root Mean Squared Error** (RMSE)
- Used to assess predictive error

## Understanding RSE

What would you say is a "good" RSE/RMSE value? What would be a "bad" one?

. . .

- You can evaluate it relative to RSE/RMSE of another model (a model with different $X$ for the same $Y$)
- RMSE is "roughly" the average error, so you can think in context if the average error is acceptable or unacceptable

## $R^2$ statistic

$$R^2 = (TSS - RSS) / (TSS)$$

$R^2$ measures percent of variability in $Y$ explained by model

- $TSS = \sum_{i = 1}^{n} (y_i - \bar{y})^2$ measures amount of variability in $Y$
- $RSS$ measures amount of variability in $Y$ unexplained by model
- $TSS - RSS$ measures amount of variability explained by the model


## $R^2$ statistic

*What* $R^2$ *be for perfectly linear data?*

*What would* $R^2$ *be for completely unlinear data?*

. . .

In practice it is context specific for what constitutes a "reasonably good" $R^2$

- consider physics versus marketing

## Cautions with $R^2$statistic

Low $R^2$ does not mean no/(or weak) relationship, simply not a linear relationship

::::{.columns}
:::{.column width="80%"}
```{r}
#| echo: false

set.seed(2343)
bad_linear_data <- data.frame(
  x = runif(100, -10, 10),
  residual = rnorm(100, mean = 0, sd = 15)
) |> 
  mutate(y = 2*x + x^2 + 1 + residual)

r_squared <- cor(bad_linear_data$x, bad_linear_data$y)^2
r_squared <- round(r_squared, 2)

ggplot(bad_linear_data, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

:::

:::{.column width="20%"}
$R^2$ = `r r_squared`
:::
::::

## Cautions with $R^2$statistic

High $R^2$ does not mean a regression line fits the data well. Another function may describe it better

::::{.columns}
:::{.column width="80%"}
```{r}
#| echo: false
bad_linear_data <- data.frame(
  x = runif(100, -10, 10),
  residual = rnorm(100, mean = 0, sd = 300)
) |> 
  mutate(y = 2*x + 4 * x^3 + 1 + residual)

r_squared <- cor(bad_linear_data$x, bad_linear_data$y)^2
r_squared <- round(r_squared, 2)

ggplot(bad_linear_data, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

:::

:::{.column width="20%"}
$R^2$ = `r r_squared`
:::
::::

## Cautions with  $R^2$

$R^2$ will always increase as $p$ increases

People developed Adjusted $R^2$, which penalizes for number of predictors BUT

- it no longer has the nice interpretation of $R^2$
- still prefers larger models



# Q4 Given a set of predictor values, what response value should we predict, and how accurate is our prediction?

##

Same as with simple linear regression!

- We can compute a prediction by plugging in values for all of our $X_j$'s into our fitted model
- We can compute a prediction interval

##

What do we predict the average test score will be for a school with a student-teacher ratio of 22, student-computer ratio of 10, and 50% qualifying for reduced-price lunch?

```{r}
my_new_data <- data.frame(
  student_teacher_ratio = 22,
  student_computer_ratio = 10,
  lunch = 50
)

predict(score_model1, newdata = my_new_data, interval = "prediction")
```

. . .

```{r}
my_pred <- predict(score_model1, newdata = my_new_data, interval = "prediction")
```


We predict their average test score to be `r round(my_pred[1], digits = 2)`, between `r round(my_pred[2], digits = 2)` and `r round(my_pred[3], digits = 2)` with 95% confidence.